---
title: "Seminar 1"
subtitle: ""
format:
  revealjs: 
    slide-number: true
    smaller: true
    scrollable: true
    incremental: true
    chalkboard: false 
    buttons: false
    preview-links: auto
    theme: default
    embed-resources: false
    code-fold: true
    code-overflow: scroll
    code-line-numbers: true
    auto-stretch: true
    html-math-method: katex
    progress: true
    bibliography: /Users/joseph/GIT/templates/bib/references.bib
---

```{r include=FALSE}
# Necessary for using dvisvgm on macOS
# See https://www.andrewheiss.com/blog/2021/08/27/tikz-knitr-html-svg-fun/
Sys.setenv(LIBGS = "/usr/local/share/ghostscript/9.53.3/lib/libgs.dylib.9.53")
font_opts <- list(dvisvgm.opts = "--font-format=woff")
```


## Goals



### By the end of this seminar you will be able to:

1.  Understand the special problems that cultural research presents for **measurement.**
2.  Understand the deeper, and prior concept of **confounding.**

### Key Concepts

1.  Measurement
2.  Validity 
3.  A confounder 


## Where Does Psychology Start?

Psychology starts with a question about how people think or behave.

## Examples of Psychological Questions?

-   How does early childhood experience affect personality and behaviour?

-   What are the effects of social media on self-esteem?

-   Why do some people believe in a God or gods and others do not?

-   Why are some people motivated to sacrifice for others?

-   Does marriage make people happy?

## Example of Psychological Questions in Cross-Cultural Psychology

-   How do early childhood experiences differ across cultures, and how do these differences impact personality and behavior development?

-   Are there cultural differences in the way social media use affects self-esteem and body image?

-   How do cultural and religious beliefs shape individual attitudes towards the concept of God or gods?

-   What are the cultural and individual factors that motivate people to engage in acts of altruism or sacrifice for others, and how do these factors vary across cultures?

-   Are there cultural differences in the factors that contribute to marital satisfaction and happiness, and how do cultural expectations and values surrounding marriage play a role?

## How might Psychological Scientists answer whether marriage makes people happy?

## *Wrong* Answers Only

-   "Ask my married parents if they are happy."

-   "Consult a palm reader."

-   Assert: "Yes marriage always makes people happy"; "No marriage can't possibly make anyone happy"

-   Intuit: "It depends on the gender of the individual. Men are always happier in marriage, while women are never happier."

-   Intuition: "It depends on the cultural background of the individuals. Couples from Western cultures are always happier in marriage, while couples from Eastern cultures are never happier."

-   Conduct a literature review of previous research on the association between marriage and happiness, including cross-cultural studies that compare different cultural attitudes and practices regarding marriage and their relationship to happiness.

-   Conduct a survey of a large and diverse sample of individuals to assess their happiness levels, as well as their marital status assess the relationship.

## Conduct a Literature Review {.incremental}

-   What would this tell us?

    -   what other researchers have found.

-   What would this not tell us?

    -   what other researchers have not found.
    -   what other researchers got wrong.

## Conduct a Survey with a Large and Diverse Sample of Individuals

-   Where to begin?

## Measurement

Definitions:

-   "Measurement is the numerical quantifcation of the attributes of an object or event, which can be used to compare with other objects or events" \[Conventional\]
-   "Measurement is the assignment of numerals to objects or events according to rules." \[Psychological\]
-   "Measurement is the process of experimentally obtaining one or more quantity values that can reasonably be attributed to a quantity." \[Metrological\] from [@briggs2021]

## "On a scale of 1-7, how happy are you?"

How might this go wrong?

-   Ambiguity: Respondents may interpret "happiness" in different ways. -e.g. Some people may equate happiness with a momentary positive emotion -e.g. Others may think of happiness as a long-term state of contentment.
-   Social Desirability Bias: Respondents may want to tell you what they think you want to hear. Others might want to frustrate you.
-   Mood: the context in which the question is asked and the respondent's current mood might affect the answer (reliability)
-   Limited range of scale
-   What do the endpoints mean?
-   Cultural Differences:
    -   in meanings
    -   in constructs themselves
    
## In Survey Research

Construct: e.g. Happiness - a subjective experience of positive emotions, such as joy, contentment, and satisfaction with life/

**Item**: A question or statement used to measure an aspect of happiness. Example: "I feel content with my life."

**Scale**: A collection of items designed to measure a construct. Example: a happiness scale might consist of items rated on a Likert-type scale, with an overall score reflecting the level of happiness being measured. Example: 

- "In general, how happy do you feel?" (rated on a scale from 1 - not at all happy to 5 - extremely happy)
- "How often do you feel positive emotions, such as joy or contentment?" (rated on a scale from 1 - very rarely or never to 5 - very often or always)
- "How satisfied are you with your life as a whole?" (rated on a scale from 1 - very dissatisfied to 5 - very satisfied)
    

## Concept of `Validity` in Psychometric Research

-   Content Validity: The degree an instrument measures what it is intended to measure (the "construct").
-   Construct Validity: Whether the construct assumed to measure is accurately defined and operationalised.
-   Criterion Validity: Whether an instrument accurately predicts performance.
-   Face Validity: whether an instrument measure what it is intended to measure, as assessed by experts.
-   Ecological Validity: whether an instrument accurately reflects real-world situations and behavior [@bandalos2018]


:::{.notes}
Metric equivalence: Factor loadings are similar across groups.
Configural equivalence: The factor structure is the same across groups in a multi-group confirmatory factor analysis.
Scalar equivalence: Values/Means are also equivalent across groups.
:::

## What else might go wrong?

Suppose we are confident in measurement, administer survey, and find an relationship.

::: {.fragment .highlight-red}
Does marriage cause happiness cross culturally?
:::


## What might go wrong?


### Happiness might *cause* marriage

```{tikz engine.opts=font_opts}
\begin{tikzpicture}
\node [draw=none, inner sep = 0] (Y) at (0, 0) {H$_1$};
\node [draw=none, inner sep = 0] (A) at (1, 0) {M$_2$};
\draw [-latex, red] (Y) to (A);
\end{tikzpicture}
```

## What else might go wrong?


### Age might a common cause of both marriage and happiness

```{tikz engine.opts=font_opts}
\begin{tikzpicture}
\node [draw=none, inner sep = 0] (A) at (0, 0) {A$_0$};
\node [draw=none, inner sep = 0] (M) at (1, 0) {M$_1$};
\node [draw=none, inner sep = 0] (H) at (2, 0) {H$_2$};
\draw [-latex, red] (A) to (M);
\draw [-latex, bend left = 30, red] (A) to  (H);
\end{tikzpicture}
```



## Many psychologists will simply "control for" income, but what if happiness and marriage cause income? 

- Simulate data in which happiness and marriage cause income
- Controlling for income makes it look as though happiness and marriage are **negatively** related.
- But we simulated data with *no* relationship!
- This is called 'collider bias', we will consider how it works in the upcoming weeks.
- Meantime, even if we measure people without error, cross-cultural studies might select people with different incomes, leading to bias **even in the absence of measurement error bias** (and **even if we do not control for income!**)

```{r}
#| echo: true

## simulate data
# reproducability
set.seed(123)
sim_fun_B <- function() {
  n <- 10000
  H <- rnorm(n, 1) # simulates Happiness,
  M <- rnorm(n, 1) #  simulates Marriage, nothing to do with happiness
  I <- rnorm(n, .2 * H + .5 * M) # simulate marriage as a function of age + happiness


  # simulate dataframe from function
  simdat_B <- data.frame(
    H = H,
    I = I,
    M = M
  )

  #  model in which marriage "predicts" happiness controlling for age
  sim_B <- lm(H ~ M + I, data = simdat_B)
  sim_B # returns output
}

# replicate 100 times
r_lm_B <- NA
r_lm_B <- replicate(100, sim_fun_B(), simplify = FALSE)

# print model results: marriage looks negatively related to happiness!
parameters::pool_parameters(r_lm_B)
```

## The problem is confounding: what can we do about it?

- Weeks 2-4: Causal Diagrams

## Our Approach: 

### External Validity

- **Definition**: the extent to which the findings of a study can be generalised to other situations, people, settings, and time periods.

- **Importance**: we want to know if our findings carry beyond the people in our study

- **Challenges**: how can we know? 

### Internal Validity

- **Definition**: the degree to which a study can demonstrate that a causal relationship exists between the 'independent' and 'dependent' variables, free of confounding.

- We will use the term 'outcome' and 'treatment' in place of independent and dependent.  This is because causality occurs in **time**.

- **Importance**: we cannot interpret the associations we recover if they do not have a causal interpretation.

- **Challenges**: internal validity requires balance in the confounders that might afffect the treatement and outcome, and observational data do not give us ballance.

- **The Special Challenges in Cross-Cultural Research**:
  - The concept of *measurement* itself has **causal** underpinnings [@vanderweele2022].
  - In cross-cultural research, we have stronger challengs from *measurement* validity, which, as we shall see, encompasses both external and internal validity.
  - We consider measurment validity within classical framework of external and internal validity
  - Ethical challenges, which science needs to consider.
  - Foremost among these ethical challenges, I believe, is doing our best to get inference right!


## The Hope

- There has been tremendous progress in the health and computer sciences in addressing problems of external and internal validity from a causal perspective
- You will be among the first to apply these methods to questions in cross-cultural psychology. 
- In doing so, much of what has been confusing to you about psychological research design, data-analysis, ultimately scientific inference will be clearer.



## What have we learned? 

::: {.fragment .highlight-red}

1.  Measurement: attaching magnitudes to features of the world.
2.  Validity: obtaining *accurate* scientific inference: a complex set of interrelated problems.
3.  Confounding; association in data is *spurious*

:::



## References

For an account of the history of measurement in psychological research, see: [@briggs2021]

For an account of key concepts and current debates in psychometrics, see: [@bandalos2018]

<!-- For an account of special measurement problems that arise in cross-cultural research, see: [@vandevijver2021] -->

For an accessible introduction to causal inference and its history see: [@pearl2018]




### Bibliography

