---
title: "Hands On Working With Quarto Manuscript"
date: "2025-MAY-13"
bibliography: /Users/joseph/GIT/templates/bib/references.bib
editor_options: 
  chunk_output_type: console
format:
  html:
    warnings: FALSE
    error: FALSE
    messages: FALSE
    code-overflow: scroll
    highlight-style: oblivion
    code-tools:
      source: true
      toggle: FALSE
html-math-method: katex
reference-location: margin
citation-location: margin
cap-location: margin
code-block-border-left: true
---

## Part 1: Quarto manuscripts

```{r}
#| include: false
#| echo: false
#read libraries

library("tinytex")
library(extrafont)
loadfonts(device = "all")

library("margot")
library("tidyverse")
```


::: {.callout-note}
**Required**
Download Quarto here: 
- Use the `prelease` version: https://quarto.org/docs/download/
**Optional**
- [@Bulbulia2024PracticalGuide] [link](https://osf.io/preprints/psyarxiv/uyg3d)
- [@hoffman2023] [link](https://arxiv.org/pdf/2304.09460.pdf)
:::

::: {.callout-important}
## Key concepts
- Writing up your manuscript

:::

::: {.callout-important}
## For the lab, download the script
- We will go through this script step-by-step.
:::



### To Do

- Download the script (from week 9) and store in our R directory: [link to analysis script ](https://raw.githubusercontent.com/go-bayes/psyc-434-2024/main/scripts/example_analysis.R)

- Download the manuscript template and store in our R directory: [link to manuscript template](https://raw.githubusercontent.com/go-bayes/psyc-434-2024/main/scripts/example-manuscript.qmd)

- Download the following `tex` file and save it to your R directory:
[latex macros](https://raw.githubusercontent.com/go-bayes/templates/main/quarto/latex.tex)


- Download the followign `tex` file and save it to your R directory:
[title preamble](https://raw.githubusercontent.com/go-bayes/templates/main/quarto/title.tex)


- Download the following `csl` file and save it to your R directory:
[csl preamble](https://raw.githubusercontent.com/go-bayes/templates/main/csl/camb-a.csl)

- Make sure you can install all libraries required of the manuscript template.
- Come to the seminar prepared to work through the analysis



## What You Will Learn

- **How to create a publication quality manuscript** 

- **How to create a workflow for references** 

- **How to import results into your manuscript** 

- **How to make graphs of your results (using) `margot`** 

- **How to report your results**

- **How to interpret your results**


### Packages

```{r}
report::cite_packages()
```




