---
title: "Hands On Working With Quarto Manuscript"
date: "2025-MAY-13"
bibliography: /Users/joseph/GIT/templates/bib/references.bib
editor_options: 
  chunk_output_type: console
format:
  html:
    warnings: FALSE
    error: FALSE
    messages: FALSE
    code-overflow: scroll
    highlight-style: Ayu
    code-tools:
      source: true
      toggle: FALSE
html-math-method: katex
reference-location: margin
citation-location: margin
cap-location: margin
code-block-border-left: true
---

## Part 1: Quarto manuscripts

```{r}
#| include: false
#| echo: false
#read libraries

library("tinytex")
library(extrafont)
loadfonts(device = "all")

library("margot")
library("tidyverse")
```


::: {.callout-note}
**Required**
Download Quarto here: 
- Use the `prelease` version: https://quarto.org/docs/download/
**Optional**
- [@Bulbulia2024PracticalGuide] [link](https://osf.io/preprints/psyarxiv/uyg3d)
- [@hoffman2023] [link](https://arxiv.org/pdf/2304.09460.pdf)
:::

::: {.callout-important}
## Key concepts
- Entering bibliographic details
- Writing up your manuscript
:::

# Code review: YOUR analysis

-  Create a new Rstudio project
-  Modify *these* scripts


::: {.callout-note-script-0}
[Download full lab scripts 0](../laboratory/lab-10/00-setup-L10.R)
:::


::: {.callout-note-script-1}
[Download full lab scripts 1](../laboratory/ ../laboratory/lab-10/01-init-L10.R)
:::

::: {.callout-note-script-2}
[Download full lab scripts 2]( ../laboratory/lab-10/02-make-wide-L10.R)
:::

::: {.callout-note-script-3}
[Download full lab scripts 3](../laboratory/lab-10/03-models-L10-v3.R)
:::




#  Quarto with `boilerplate`

::: {.callout-quarto-manuscript-template}
[Download Quarto Template](../quarto/initial_quarto_document.qmd)
:::


## Script 0: Set up your library/

Note -- rerun this script if you are missing updated templates, as they have been updated 23 May 2025

```{r}
#| include: true
#| eval: false
#| file: ../laboratory/lab-11/00-boilerplate-setup.R
```

::: {.callout-note}
Type `y` when R asks whether to overwrite the boilerplate files; this replaces them with the latest versions.
Press `n` (or just hit Enter) to keep what you already have.

Your own edits—such as custom rows in the measures table -- are stored elsewhere, so they will not be overwritten. One tap, no tears.

Take-away: `y` refreshes the templates; `n` preserves the status quo.
:::

This is how the the printout in your R console should look (depending on how recently you've run your script):

```r
> # save imported data -------------------------------------------------------
> boilerplate_save(
+   student_unified_db,
+   data_path     = path_data,
+   create_backup = FALSE
+ )
ℹ preparing to save 6 databases
ℹ processing measures database (1/6)
ℹ no changes detected in measures database
ℹ saving measures database to /Users/your_machine_path-/example_boilerplate_data/measures_db.rds
✔ saved measures database
ℹ processing methods database (2/6)
ℹ 1 new entries will be added:
ℹ   + student_target_population
! 2 entries will be removed:
!   - target_population
!   - sample
Save methods database with these changes? [y/n]: y
ℹ saving methods database to /Users/joseph/GIT/psych-434-2025/example_boilerplate_data/methods_db.rds
✔ saved methods database
ℹ processing results database (3/6)
ℹ no changes detected in results database
ℹ saving results database to /Users/joseph/GIT/psych-434-2025/example_boilerplate_data/results_db.rds
✔ saved results database
ℹ processing discussion database (4/6)
ℹ 2 existing entries will be modified:
ℹ   ~ student_authors_statement
ℹ   ~ student_ethics
Save discussion database with these changes? [y/n]: y
ℹ saving discussion database to /Users/joseph/GIT/psych-434-2025/example_boilerplate_data/discussion_db.rds
✔ saved discussion database
ℹ processing appendix database (5/6)
ℹ no changes detected in appendix database
ℹ saving appendix database to /Users/joseph/GIT/psych-434-2025/example_boilerplate_data/appendix_db.rds
✔ saved appendix database
ℹ processing template database (6/6)
ℹ no changes detected in template database
ℹ saving template database to /Users/joseph/GIT/psych-434-2025/example_boilerplate_data/template_db.rds
✔ saved template database
✔ successfully saved all 6 databases
```

## Script 1: Adding/Revising Measures

Note -- rerun this script if you are missing updated templates, as they have been updated 23 May 2025

```{r}
#| include: true
#| eval: false
#| file: ../laboratory/lab-11/01-add-revise-measures.R

```

::: {.callout-note}
- remember to type 'y' to save changes
:::


When re-running your scripts, the printout should allow you to save to your working database folder. 
When saving tick 'y'

```r
> # save a copy to avoid overwriting original --------------------------------
> boilerplate_save(
+   unified_db,
+   data_path   = path_final,
+   output_file = "unified_db"
+ )
ℹ preparing to save 6 databases
ℹ processing measures database (1/6)
ℹ no changes detected in measures database
ℹ saving measures database to /Users/joseph/GIT/psych-434-2025/final_boilerplate_data/measures_db.rds
✔ saved measures database
ℹ processing methods database (2/6)
ℹ 1 new entries will be added:
ℹ   + student_target_population
! 2 entries will be removed:
!   - target_population
!   - sample
Save methods database with these changes? [y/n]: y
ℹ created backup at: /Users/your_path/final_boilerplate_data/methods_db.rds.20250523_131930.bak
ℹ saving methods database to  /Users/your_path/final_boilerplate_data/methods_db.rds
✔ saved methods database
ℹ processing results database (3/6)
ℹ no changes detected in results database
ℹ saving results database to  /Users/your_path/final_boilerplate_data/results_db.rds
✔ saved results database
ℹ processing discussion database (4/6)
ℹ 2 existing entries will be modified:
ℹ   ~ student_authors_statement
ℹ   ~ student_ethics
Save discussion database with these changes? [y/n]: y
ℹ created backup at: /Users/joseph/GIT/psych-434-2025/final_boilerplate_data/discussion_db.rds.20250523_131933.bak
ℹ saving discussion database to /Users/joseph/GIT/psych-434-2025/final_boilerplate_data/discussion_db.rds
✔ saved discussion database
ℹ processing appendix database (5/6)
ℹ no changes detected in appendix database
ℹ saving appendix database to /Users/joseph/GIT/psych-434-2025/final_boilerplate_data/appendix_db.rds
✔ saved appendix database
ℹ processing template database (6/6)
ℹ no changes detected in template database
ℹ saving template database to /Users/joseph/GIT/psych-434-2025/final_boilerplate_data/template_db.rds
✔ saved template database
✔ successfully saved all 6 databases
```



## Script 3: Setting Up Your Manuscript Document

::: {.callout-note}
- save this document as a `.qmd` file in a directory called 'quarto' (in your rstudio project)
- - to render a document click 'Render'
:::





```{r}
#| include: true
#| eval: false
#| file: ../quarto/initial_quarto_document.qmd

```

#### Pro Tip

You can adjust figure by changing:

`#| fig-height: 16` and `#| fig-width: 9`


```r
#| label: fig-policy-6
#| fig-cap: "Decision Tree: {glued_policy_names_6}"
#| eval: true
#| fig-height: 16 # <- change here
#| fig-width: 9 # <- change here

policy_plots[[6]]

```



## What You Have Learned 

- **How to create a publication quality manuscript** 

- **How to create a workflow for references** 

- **How to import results into your manuscript** 

- **How to make graphs of your results (using) `margot`** 

- **How to report your results**

- **How to interpret your results**

## Frequently Asked Questions

### Do I need to use 'quarto' for making my report?

No. The suggested workflow should make your life *easier*. If, after a fair trial -- *and* after asking us for help -- `quarto` still feels like extra friction, switch to whatever you like (`rmarkdown`, plain `R` scripts, or even Word). The science matters more than the wrapper. 

👉 Quick-start: <https://quarto.org/docs/get-started/>

---

### I cannot save or retrieve files from folders

When hunting path errors, ask yourself:

- **Packages** – have I updated everything, especially **`margot`** and **`here`**?  
- **Clean slate** – did I restart R (⌘⇧F10) *and* rerun the data-prep scripts from the top?  
- **Project root** – am I inside the correct **RStudio Project**? Check with `here::here()`.  
- **Search** – have I used *Edit → Find in Files* (⌘⇧F / Ctrl⇧F) to locate the object or path?  
- **Data really there?** – do the objects exist in my data frame?

```r
# waves where `variable_name` was measured
with(df_nz_long, table(variable_name, wave))
```

- **Permissions / cloud sync** – is the save folder writable *and* local? Dropbox/OneDrive may off-load files; right-click and choose *Make available offline*.

---

### I cannot run 'margot_causal_forest_parallel()'

- **Fallback** – does `margot_causal_forest()` work? It is slower but prints progress.  
- **Formula richness** – causal forests need many covariates [@grf2024]. Include the full demographic block.  
- **Resources** – parallel forests spawn one worker per core. Keep some RAM free.  
- **Reduce load** – try fewer trees or a smaller variable pool:

```r
grf_defaults <- list(
  seed             = 123,
  stabilize.splits = TRUE,
  num.trees        = 1000,  # fewer trees
  top_n_vars       = 10     # smaller policy-tree search space
)
```

---

### I cannot run 'margot_flip_forests_parallel()'

First test the serial version `margot_flip_forests()`. All parallel caveats above apply.

---

### My R code is not working

- Update packages (`pak::pak()` is handy).  
- Re-download the example code and run from the top.  
- Read the full error and traceback (`rlang::last_trace()`).  
- Compare with the example: what is different in *your* data or model?

---

### I do not know how 'margot' package functions work

See the indexed reference: <https://go-bayes.github.io/margot/reference/index.html>

---

### What is the recommended workflow for investigating heterogeneity?

1. **State your causal question.** Statistics cannot rescue an ill-posed question.  
2. Estimate the **ATE** for the target population.  
3. Compute **CATEs** and evaluate gain at 20 % and 50 % budget constraints with the **Qini** metric.  
4. For outcomes with convincing heterogeneity, fit shallow **policy trees** (depth = 2) to obtain transparent rules.  
5. Optionally, compare pre-specified subgroups directly.  
6. Report **RATE–AUTOC** and **RATE–Qini** tables in an online supplement.  

*Take-away:* start from a clean, updated project, read errors carefully, and ensure each step answers *your* research question.

---

### I cannot create a 'quarto' document

- Save the file with a `.qmd` suffix **inside** the `quarto` folder.  
- Ensure the boiler-plate directory structure exists (run Script 0 and Script 1).  
- Run the initial code block line-by-line before hitting **Render**.  
- Define all `global_vars` required by the boiler-plate.  
- If you are *not* flipping outcomes, stub the flip variables:

```r
flipped_names <- ""
flip_outcomes <- ""
flipped_list  <- ""
```

::: {.callout-note}
If you are not flipping forests, comment out or edit any boiler-plate code that assumes flips.
:::

---

### I cannot render the 'quarto' document as PDF

Install **TinyTeX** once:

```r
install.packages("tinytex")
tinytex::install_tinytex()
```

Then, in your YAML front-matter, comment out unwanted formats. For PDF-only output:

```yaml
format:
  # docx:
  #   default: false
  pdf:
    pdf-engine: lualatex
```

---

### Is it plagiarism to use the boiler-plate outputs verbatim?

No—provided you clearly cite the source. The templates come from the  EPIC (Bulbulia) Lab and New Zealand Attitudes and Values Study. Tailor your text to match your study and give credit. To edit, set `eval: false`, render, and copy-paste the plain text blocks.

---

### I do not have a boiler-plate template used in the example script

Run **Script 0** and **Script 1** from the start; they refresh every template.
  
  
```{r}
report::cite_packages()
```




