---
title: "Hands On Working With Quarto Manuscript"
date: "2025-MAY-13"
bibliography: /Users/joseph/GIT/templates/bib/references.bib
editor_options: 
  chunk_output_type: console
format:
  html:
    warnings: FALSE
    error: FALSE
    messages: FALSE
    code-overflow: scroll
    highlight-style: Ayu
    code-tools:
      source: true
      toggle: FALSE
html-math-method: katex
reference-location: margin
citation-location: margin
cap-location: margin
code-block-border-left: true
---

## Part 1: Quarto manuscripts

```{r}
#| include: false
#| echo: false
#read libraries

library("tinytex")
library(extrafont)
loadfonts(device = "all")

library("margot")
library("tidyverse")
```


::: {.callout-note}
**Required**
Download Quarto here: 
- Use the `prelease` version: https://quarto.org/docs/download/
**Optional**
- [@Bulbulia2024PracticalGuide] [link](https://osf.io/preprints/psyarxiv/uyg3d)
- [@hoffman2023] [link](https://arxiv.org/pdf/2304.09460.pdf)
:::

::: {.callout-important}
## Key concepts
- Entering bibliographic details
- Writing up your manuscript
:::

# Code review: YOUR analysis

-  Create a new Rstudio project
-  Modify *these* scripts


::: {.callout-note-script-0}
[Download full lab scripts 0](../laboratory/lab-10/00-setup-L10.R)
:::


::: {.callout-note-script-1}
[Download full lab scripts 1](../laboratory/ ../laboratory/lab-10/01-init-L10.R)
:::

::: {.callout-note-script-2}
[Download full lab scripts 2]( ../laboratory/lab-10/02-make-wide-L10.R)
:::

::: {.callout-note-script-3}
[Download full lab scripts 3](../laboratory/lab-10/03-models-L10-v3.R)
:::




#  Quarto with `boilerplate`

::: {.callout-quarto-manuscript-template}
[Download Quarto Template](../quarto/initial_quarto_document.qmd)
:::


## Script 0: Set up your library/


```{r}
#| include: true
#| eval: false
#| file: ../laboratory/lab-11/00-boilerplate-setup.R
```

## Script 1: Adding/Revising Measures

```{r}
#| include: true
#| eval: false
#| file: ../laboratory/lab-11/01-add-revise-measures.R

```


## Script 3: Setting Up Your Manuscript Document

Note -- save this as a '.qmd' file

```{r}
#| include: true
#| eval: false
#| file: ../quarto/initial_quarto_document.qmd

```



## What You Will Learn

- **How to create a publication quality manuscript** 

- **How to create a workflow for references** 

- **How to import results into your manuscript** 

- **How to make graphs of your results (using) `margot`** 

- **How to report your results**

- **How to interpret your results**


### Packages

```{r}
report::cite_packages()
```




